\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-05-03 doc/publications/plos16/nm.tex
Short abstract

% Please keep the abstract below 300 words
\section{Abstract}
Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
novel candidates.

\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
real-time classification of population-scale patient cohorts in ``patients like mine'' scenarios, as well as performing
GWAS analysis on large unfiltered whole genome sequencing cohorts.

>>>>

% Please keep the abstract below 300 words
\section{Abstract}
%Modern data analyses are performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
%genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
%differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
%compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
%data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
%they scale well with samples ($n$) but not with features ($p$, variants).
Modern genomic sequencing studies have seen increasingly large cohort in the past decade. This lead to genomic sequencing datasets with not only bigger sample sizes,
but also, more crucially, exponentially growing variant space. Traditional data analysis approaches are unable to cater such kind of big and wide datasets. Even the machine
learning library in the highly parallelised Spark computing engine suffer from this ``the curse of dimensionality'.
We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. With public genome sequencing and SNP array datasets, we demonstrate the utilities of \cursedforest 
for accurately predicting sample ethnicity and identifying known and novel variants associated with the phenotype of interest.
%We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
%profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
%was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
%more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
%dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
%novel candidates. 
\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
real-time classification of population-scale patient cohorts in ``patients like mine'' scenarios, as well as performing
GWAS-style analysis on large unfiltered whole genome sequencing cohorts.
